State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-04-14 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
updates and edits to paper

\title[State and Parameter Estimation for a Spatio-Temporal Neural Fields]{State and Parameter Estimation for a Spatio-Temporal Neural Fields}

@@

The firing rate of the presynaptic neurons is related to the postsynaptic membrane potential by the sigmoidal activation function
\begin{equation}\label{ActivationFunction}
	f\left( v\left( \mathbf{r}', t \right) \right) = \frac{f_{max}\left(\mathbf{r}\right)}{1 + \exp \left( \varsigma \left( v_0\left( \mathbf{r} \right) - v\left(\mathbf{r}',t\right) \right) \right)}.
\end{equation}
%Even though we will model $v_0$ and $\nu_0$ as being homogeneous across the cortical sheet we will express it as being spatially dependant allowing for generality at this point. 
By substituting \ref{RateBasedInteractions} into \ref{SpikesToPotential}  we get the spatiotemporal model
\begin{equation}\label{FullDoubleIntModel}
	v\left(\mathbf{r},t\right) = \int_{-\infty}^t h\left(t - t'\right) \left(\int_\Omega   w\left(\mathbf{r},\mathbf{r}'\right) f\left( v\left( \mathbf{r}',t \right)\right)d\mathbf{r}' + u(\mathbf{r},t) \right)dt'.
\end{equation}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function
\begin{equation}\label{GreensFuncDef}
	Lh\left( t \right) = \delta \left( t \right),
\end{equation}
where $L=\frac{d}{dt} + \zeta$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function giving 
\begin{equation}\label{FinalFormContinuous}
	\frac{dv\left( \mathbf{r},t \right)}{dt} + \zeta v\left( \mathbf{r},t \right) = \int_\Omega  {w\left( \mathbf{r},\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)dr'} + u\left(\mathbf{r},t\right).
\end{equation}
To arrive at the integro-difference equation (IDE) form of the model we discretize time using a first-order Euler method (see Appendix~\ref{Time Discretization}) giving
\begin{equation}\label{DiscreteTimeModel}
	v_{t+1}\left(\mathbf{r}\right) = \xi v_t\left(\mathbf{r}\right) + T \int_\Omega { w\left(\mathbf{r}-\mathbf{r}'\right) f\left(v_t\left(\mathbf{r}'\right)\right) d\mathbf{r}'} + T u_t\left(\mathbf{r}\right) + e_t\left(\mathbf{r}\right),
\end{equation} 
where $T$ is the time step, $\xi = 1-\zeta T$ and is an iid spatially colored, temporally white disturbance term with the covariance 
\begin{equation}
    cov(\mathbf{e}_t,\mathbf{e}_{t+\tau}) = \left\{ \begin{array}{*{20}{c}}
    \gamma\left(\mathbf{r}-\mathbf{r'}\right), & \tau=0 \\
    0 & \mathrm{otherwise}. \\
    \end{array} \right.
\end{equation}
This term was added to drive the system and account for uncertainty in the model. To generate iEEG/LFP data we used the output function 
\begin{equation}
	\mathbf{y}_t = \int_{\Omega}{m\left(\mathbf{r}_n-\mathbf{r}'\right)v_t\left(\mathbf{r}'\right)d\mathbf{r}'} + \boldsymbol{\varepsilon}_t,
\end{equation}
where $\mathbf{r}_n$ defines the location of the sensors in the field where $n=1,...,N$ indexes the sensors and $\varepsilon_t \sim \mathcal{N}\left(0,\Sigma{\varepsilon}\right)$. The output kernel $m(\mathbf{r}-\mathbf{r}')$ governs the geometry of pick-up range of sensors where
\begin{equation}
	m\left(\mathbf{r}-\mathbf{r}'\right) = \exp{\left(-\frac{(\mathbf{r}-\mathbf{r}')^\top(\mathbf{r}-\mathbf{r}')}{\sigma_m^2}\right)}.
\end{equation}

@@

\boldsymbol{\Sigma}_e=\boldsymbol{\Gamma}^{-1}\int\limits_{\Omega}\int\limits_{\Omega}\boldsymbol{\phi}\left(\mathbf r\right) \gamma\left(\mathbf{r}- \mathbf{r}' \right)\boldsymbol{\phi}\left(\mathbf{r}'\right)^{\top}d\mathbf{r}' d\mathbf{r}\boldsymbol{\Gamma}^{- \top} 

@@

    \mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\varepsilon}_t \\

@@

\mathbf{C} = \left[\begin{array}{{cccc}}
\int_{\Omega}m(\mathbf{r}_1 - \mathbf{r}')\boldsymbol{\phi}_1(\mathbf{r}')d\mathbf{r}' & \int_{\Omega} m(\mathbf{r}_1 - \mathbf r')\boldsymbol \phi_2(\mathbf r')d\mathbf r' & \dots
&\int_{\Omega}m(\mathbf r_1 - \mathbf r')\boldsymbol \phi_n(\mathbf
r')d\mathbf r' \\
\int_{\Omega}m(\mathbf r_2 - \mathbf r')\boldsymbol \phi_1(\mathbf
r')d\mathbf r'&\int_{\Omega}m(\mathbf r_2 - \mathbf r')\boldsymbol
\phi_2(\mathbf r')d\mathbf r'& \dots &\int_{\Omega}m(\mathbf r_2 -
\mathbf r')\boldsymbol \phi_n(\mathbf r')d\mathbf r'\\
\vdots&\vdots&\ddots&\vdots\\
\int_{\Omega}m(\mathbf r_{n_y} - \mathbf r')\boldsymbol \phi_1(\mathbf
r')d\mathbf r'&\int_{\Omega}m(\mathbf r_{n_y} - \mathbf r')\boldsymbol
\phi_2(\mathbf r')d\mathbf r'& \dots &\int_{\Omega}m(\mathbf r_{n_y} -
\mathbf r')\boldsymbol \phi_n(\mathbf r')d\mathbf r'
\end{array}\right],

@@

and $m(\mathbf r_n - \mathbf{r}')$ is the observation kernel.

@@

Considering the spectral representation of the dynamic field at time $t$ denoted by $W_t(\boldsymbol{\nu})$ (\todo{can this be $V_t(\boldsymbol{\nu})$}), multi-dimensional sampling and reconstruction theory dictates that for and accurate representation, $W_t(\boldsymbol{\nu})$ need to be spatially bandlimited. Such a condition is usually satisfied due to the bandlimiting effect of most practical kernels and the spatially colored covariance functions \todo{~cite\{some eeg literature here\}}. In addition, the Gaussian pick-up of the sensors provides a low-pass filtering effect, further limiting the effective bandwidth of the field. Thus the spectral response $W_t(\boldsymbol{\nu})$ is usually band-limited with $W_t(\boldsymbol{\nu}) \approx 0 ~ \forall \boldsymbol{\nu} > \boldsymbol{\nu}_c$, where $\boldsymbol{\nu}_c$ is a cutoff frequency. Note that for the 3-dimensional, homogeneous and isotopic fields being considered $\boldsymbol{\nu}_c = [\nu_c ~ \nu_c ~ \nu_c]^\top$. Given such a band-limited field and based on Shannon's sampling theorem, the distance between adjacent sensors, $\Delta_s$, needs to satisfy
\begin{equation}
\Delta_s = \frac{1}{2\rho\nu_c},
\end{equation}
where $\rho \in \mathbb{R} \ge 1$ is an oversampling parameter. Satisfying this condition is required for the hidden dynamic field $w_t(\mathbf{r})$ to be precisely inferred from the sampled observations, $\mathbf{y}(t)$. If the condition is not satisfied, spatial aliasing will occur.

There are two options available to alleviate problems with aliasing. The first is to increase the spatial sampling frequency (using a higher resolution of sensors), and the second is to use wider sensors, which act like an anti-aliasing filters. The first option is more desirable, since more information is retained. However, if sensors are modelled as having a Gaussian pickup range than the effective filter will not have a sharp cut-off and high frequency components will still remain and mild aliasing will still occur. In practice, the bandwidth of the cortex can not be estimated by traditional electrophysiological measurements due to spatial limitations. It may be estimated using other modalities with higher spatial resolution such as, fMRI, NIRS or other optical imaging techniques. Errors associated with mild aliasing result in distortions in the high frequency components of the field dynamics. However, after applying the basis function decomposition, high frequency component contribute negligibly to the inverse problem and are therefore not a major problem for the state and parameter estimation~\cite{Sanner1992}. This is further illustrated our results.
$\mathcal{W}$
Similar considerations need to be made regarding the representation of the dynamic field $w_t(\mathbf{r})$ by its basis functions decompositions. The minimum distance between adjacent basis functions also needs to satisfy Shannon's sampling theorem. Thus, the minimum distance between basis must satisfy
\begin{equation}
\Delta_b < \frac{1}{2\nu_c}.
\end{equation}
For the Gaussian basis functions $\phi(\mathbf{r})$ being considered, the basis width parameter $\sigma^2$ can also be inferred from spectral considerations \cite{Sanner1992,Scerri2009}. The Fourier transform of a Gaussian is another Gaussian with width $\sigma^2_{\nu}$ given by
\begin{equation}
\sigma^2_{\nu} = \frac{1}{\pi^2\sigma^2}.
\end{equation}
Thus to ensure that the basis functions can represent fields with frequency content up to $\boldsymbol{\nu}_c$, $\sigma_{\nu}$ should be set with the 3~dB frequency for the basis functions equal to $\boldsymbol{\nu}_c$, thus obtained $\sigma^2_{\nu}$ and therefore $\sigma^2$.

The dynamic field must be spatially oversampled by using a larger number of sensor then imposed by frequency considerations, to use frequency analysis to determine the minimum distance and width of the basis functions. Following this, the distance between basis functions is usually larger then the distance between sensors. \todo{However this is not the case with this work.} For a spatially homogeneous isotropic field the basis functions can be placed on a regular square grid. Following this, the knowledge of the distance between basis functions directly implies the total number of basis functions required.

>>>>

\title[State and Parameter Estimation for Spatio-Temporal Neural Fields]{State and Parameter Estimation for Spatio-Temporal Neural Fields}

@@

The firing rate of the presynaptic neurons is related to the postsynaptic membrane potential by the sigmoidal activation function
\begin{equation}\label{ActivationFunction}
	f\left( v\left( \mathbf{r}', t \right) \right) = \frac{f_{max}\left(\mathbf{r}\right)}{1 + \exp \left( \varsigma \left( v_0\left( \mathbf{r} \right) - v\left(\mathbf{r}',t\right) \right) \right)}.
\end{equation}
%Even though we will model $v_0$ and $\nu_0$ as being homogeneous across the cortical sheet we will express it as being spatially dependant allowing for generality at this point. 
By substituting \ref{RateBasedInteractions} into \ref{SpikesToPotential}  we get the spatiotemporal model
\begin{equation}\label{FullDoubleIntModel}
	v\left(\mathbf{r},t\right) = \int_{-\infty}^t h\left(t - t'\right) \left(\int_\Omega   w\left(\mathbf{r},\mathbf{r}'\right) f\left( v\left( \mathbf{r}',t \right)\right)d\mathbf{r}' + u(\mathbf{r},t) \right)dt'.
\end{equation}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function
\begin{equation}\label{GreensFuncDef}
	Dh\left( t \right) = \delta \left( t \right),
\end{equation}
where $D=\frac{d}{dt} + \zeta$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function giving 
\begin{equation}\label{FinalFormContinuous}
	\frac{dv\left( \mathbf{r},t \right)}{dt} + \zeta v\left( \mathbf{r},t \right) = \int_\Omega  {w\left( \mathbf{r},\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)dr'} + u\left(\mathbf{r},t\right).
\end{equation}
To arrive at the integro-difference equation (IDE) form of the model we discretize time using a first-order Euler method (see Appendix~\ref{Time Discretization}) giving
\begin{equation}\label{DiscreteTimeModel}
	v_{t+1}\left(\mathbf{r}\right) = \xi v_t\left(\mathbf{r}\right) + T \int_\Omega { w\left(\mathbf{r}-\mathbf{r}'\right) f\left(v_t\left(\mathbf{r}'\right)\right) d\mathbf{r}'} + T u_t\left(\mathbf{r}\right) + e_t\left(\mathbf{r}\right),
\end{equation} 
where $T$ is the time step, $\xi = 1-\zeta T$ and $e_t\left(\mathbf{r}\right)$ is an iid spatially coloured, temporally white disturbance term with the covariance 
\begin{equation}
    cov(\mathbf{e}_t,\mathbf{e}_{t+\tau}) = \left\{ \begin{array}{*{20}{c}}
    \gamma\left(\mathbf{r}-\mathbf{r'}\right), & \tau=0 \\
    0 & \mathrm{otherwise}. \\
    \end{array} \right.
\end{equation}
This term was added to drive the system and account for uncertainty in the model. To generate iEEG/LFP data we used the output function 
\begin{equation}
	\mathbf{y}_t = \int_{\Omega}{m\left(\mathbf{r}_n-\mathbf{r}'\right)v_t\left(\mathbf{r}'\right)d\mathbf{r}'} + \boldsymbol{\varepsilon}_t,
\end{equation}
where $\mathbf{r}_n$ defines the location of the sensors in the field where $n=1,...,N$ indexes the sensors and $\boldsymbol{\varepsilon}_t \sim \mathcal{N}\left(0,\Sigma{\varepsilon}\right)$. The output kernel $m(\mathbf{r}-\mathbf{r}')$ governs the geometry of pick-up range of sensors where
\begin{equation}
	m\left(\mathbf{r}-\mathbf{r}'\right) = \exp{\left(-\frac{(\mathbf{r}-\mathbf{r}')^\top(\mathbf{r}-\mathbf{r}')}{\sigma_m^2}\right)}.
\end{equation}

@@

\boldsymbol{\Sigma}_e=\boldsymbol{\Gamma}^{-1}\int_{\Omega}\int_{\Omega}\boldsymbol{\phi}\left(\mathbf r\right) \gamma\left(\mathbf{r}- \mathbf{r}' \right)\boldsymbol{\phi}\left(\mathbf{r}'\right)^{\top}d\mathbf{r}' d\mathbf{r}\boldsymbol{\Gamma}^{- \top} 

@@

    \mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\varepsilon}_t

@@

	\mathbf{C} = \left[\begin{array}{{ccc}}
	c_{1,1} & \dots & c_{1,L} \\
	\vdots & \ddots & \vdots \\
	c_{n_y,L} & \dots & c_{n_y,L} \end{array}\right]

@@

where $L$ is the number of basis functions and
	c_{i,j} = \int_{\Omega}m(\mathbf{r}_i - \mathbf{r}')\boldsymbol{\phi}_j(\mathbf{r}')d\mathbf{r}'.
% \begin{equation}
% \mathbf{C} = \left[\begin{array}{{cccc}}
% \int_{\Omega}m(\mathbf{r}_1 - \mathbf{r}')\boldsymbol{\phi}_1(\mathbf{r}')d\mathbf{r}' & \int_{\Omega} m(\mathbf{r}_1 - \mathbf r')\boldsymbol \phi_2(\mathbf r')d\mathbf r' & \dots
% &\int_{\Omega}m(\mathbf r_1 - \mathbf r')\boldsymbol \phi_n(\mathbf
% r')d\mathbf r' \\
% \int_{\Omega}m(\mathbf r_2 - \mathbf r')\boldsymbol \phi_1(\mathbf
% r')d\mathbf r'&\int_{\Omega}m(\mathbf r_2 - \mathbf r')\boldsymbol
% \phi_2(\mathbf r')d\mathbf r'& \dots &\int_{\Omega}m(\mathbf r_2 -
% \mathbf r')\boldsymbol \phi_n(\mathbf r')d\mathbf r'\\
% \vdots&\vdots&\ddots&\vdots\\
% \int_{\Omega}m(\mathbf r_{n_y} - \mathbf r')\boldsymbol \phi_1(\mathbf
% r')d\mathbf r'&\int_{\Omega}m(\mathbf r_{n_y} - \mathbf r')\boldsymbol
% \phi_2(\mathbf r')d\mathbf r'& \dots &\int_{\Omega}m(\mathbf r_{n_y} -
% \mathbf r')\boldsymbol \phi_n(\mathbf r')d\mathbf r'
% \end{array}\right],
% \end{equation}
% and $m(\mathbf r_n - \mathbf{r}')$ is the observation kernel.

@@

Considering the spectral representation of the dynamic field at time $t$ denoted by $V_t(\boldsymbol{\nu})$, multi-dimensional sampling and reconstruction theory dictates that for and accurate representation, $V_t(\boldsymbol{\nu})$ need to be spatially bandlimited. Such a condition is usually satisfied due to the bandlimiting effect of most practical kernels and the spatially colored covariance functions \todo{~cite\{some eeg literature here\}}. In addition, the Gaussian pick-up of the sensors provides a low-pass filtering effect, further limiting the effective bandwidth of the field. Thus the spectral response $V_t(\boldsymbol{\nu})$ is usually band-limited with $V_t(\boldsymbol{\nu}) \approx 0 ~ \forall \boldsymbol{\nu} > \boldsymbol{\nu}_c$, where $\boldsymbol{\nu}_c$ is a cutoff frequency. Note that for the 3-dimensional, homogeneous and isotopic fields being considered $\boldsymbol{\nu}_c = [\nu_c ~ \nu_c ~ \nu_c]^\top$. Given such a band-limited field and based on Shannon's sampling theorem, the distance between adjacent sensors, $\Delta_s$, needs to satisfy
\begin{equation}
\Delta_s = \frac{1}{2\rho\nu_c},
\end{equation}
where $\rho \in \mathbb{R} \ge 1$ is an oversampling parameter. Satisfying this condition is required for the hidden dynamic field $v_t(\mathbf{r})$ to be precisely inferred from the sampled observations, $\mathbf{y}(t)$. If the condition is not satisfied, spatial aliasing will occur.

There are two options available to alleviate problems with aliasing. The first is to increase the spatial sampling frequency (using a higher resolution of sensors), and the second is to use wider sensors, which act like an anti-aliasing filters. The first option is more desirable, since more information is retained. However, if sensors are modelled as having a Gaussian pickup range than the effective filter will not have a sharp cut-off and high frequency components will still remain and mild aliasing will still occur. In practice, the bandwidth of the cortex can not be estimated by traditional electrophysiological measurements due to spatial limitations. It may be estimated using other modalities with higher spatial resolution such as, fMRI, NIRS or other optical imaging techniques. Errors associated with mild aliasing result in distortions in the high frequency components of the field dynamics. However, after applying the basis function decomposition, high frequency component contribute negligibly to the inverse problem and are therefore not a major problem for the state and parameter estimation~\cite{Sanner1992}. This is further illustrated our results.

Similar considerations need to be made regarding the representation of the dynamic field $w_t(\mathbf{r})$ by its basis functions decompositions. The minimum distance between adjacent basis functions also needs to satisfy Shannon's sampling theorem. Thus, the minimum distance between basis must satisfy
\begin{equation}
\Delta_b < \frac{1}{2\nu_c}.
\end{equation}
For the Gaussian basis functions $\phi(\mathbf{r})$ being considered, the basis width parameter $\sigma^2$ can also be inferred from spectral considerations \cite{Sanner1992,Scerri2009}. The Fourier transform of a Gaussian is another Gaussian with width $\sigma^2_{\nu}$ given by
\begin{equation}
\sigma^2_{\nu} = \frac{1}{\pi^2\sigma^2}.
\end{equation}
Thus to ensure that the basis functions can represent fields with frequency content up to $\boldsymbol{\nu}_c$, $\sigma_{\nu}$ should be set with the 3~dB frequency for the basis functions equal to $\boldsymbol{\nu}_c$, thus obtained $\sigma^2_{\nu}$ and therefore $\sigma^2$.

The dynamic field must be spatially oversampled by using a larger number of sensor then imposed by frequency considerations, to use frequency analysis to determine the minimum distance and width of the basis functions. Following this, the distance between basis functions is usually larger then the distance between sensors, although this is not a requirement for using this estimation framework. For a spatially homogeneous isotropic field the basis functions can be placed on a regular square grid. Following this, the knowledge of the distance between basis functions directly implies the total number of basis functions required.
